\RequirePackage[l2tabu, orthodox]{nag}
%\documentclass[a4paper,12pt]{article}
\documentclass[man,floatsintext]{apa6} % man, doc, jou

\usepackage{amsmath}
% \usepackage[a4paper]{geometry}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{apacite}
% \usepackage{natbib}
%\usepackage{newclude} % \include* to include without page break
%\usepackage[colorlinks=false, pdfborder={0 0 0}]{hyperref}
\usepackage{cleveref} % \cref instead of \ref, automatic "Section" etc.
\usepackage{eurosym}

\title{Predicting reaction time using oscillations in the beta band}
\shorttitle{Predicting reaction time with beta activity}
\author{Max Hollmann}
\affiliation{University of Groningen}
\date{\today}
\authornote{Thanks to Hedderik van Rijn, Mark Span, Udo Boehm, and Jacob Jolij.}

\abstract{
  Abstract...
}
\keywords{beta power, interval timing, reaction time}

\begin{document}
\maketitle
% TODO no newline between \maketitle and Intro.

%% Hedderik: Slow potentials...
% History of area (mostly specific stuff (CNV in combination with timing))
% Explain terms (What's CNV?)
% Prior research on topic
% Explanations/mechanisms given by prior papers
% New explanation/mechanism from other research
%

% Hedderik in the method:
% The Method section is pretty technical, which is perfectly fine.
% However, this does mean that you will need to explain the rationale of
% the setup of the experiment (i.e., waiting period, then cue and as
% fast as possible reaction) in the introduction

%%% Gambit: Time perception, timing, etc
%%% Neural mechanism of timing?
%%% CNA/CNV
% CNA: climbing neural activity
% CNV: negative buildup of EEG "when subject is expecting an event"
%   proposed to reflect the accumulation part of the pacemaker-accumulator models of timing
%   Macar et al 1999: larger cnv for longer produced intervals (also for longer judged intervals)
%   Not reproduced by Kononowitz & Van Rijn 2011
%%% Beta in tapping experiments (Bartolo, Joundi)
% Beta power at beginning of interval predicts length of produced interval in synchronization-continuation tapping experiment
%%% Beta in interval production (Kononowitz & van Rijn in prep)
% Beta power at beginning of interval predicts length of produced interval
%%% Interval timing in reaction time (Grosjean? or something similar)
% Interval timing plays a role in reaction time.
%%% (Internal/external cues)

The passage of time is one of the most basic facts of our world,
making timing one of the most important functions of the mind.  Thus,
a lot of scientific effort is spent on discovering its underlying
mechanisms.

% CNV -> interval length
One line of research in this area is looking at connections between
electrophysiological measures from the brain and interval timing.  For
example, \citeA{macar_supplementary_1999} found a positive correlation
between the amplitude of a negative cortical potential during a
self-produced interval and the duration of the same interval.
However, in a replication of this experiment
\citeA{kononowicz_slow_2011} failed to find the same effect.

% Beta -> interval length
% tapping -> off/onset same -> no causality
More recently, the power of neuronal oscillations in the beta band was
found to be another measure that is predictive of the length of
produced intervals.

\citeA{jenkinson_new_2011} suggested that beta power in the basal
ganglia provides an index of the likelihood that a voluntary movement
will be required, with a low beta power indicating a high probability.
Indeed, \citeA{joundi_persistent_2013} found that in a task where
participants were instructed to tap along with an external rhythm,
beta power was lower during the tapping.  Furthermore, they found that
with high tapping rates, beta activity could not fully resynchronize
between taps.

Along similar lines, \citeA{bartolo_information_2014} used beta power
to predict the interval length in a synchronization-continuation
tapping task performed on monkeys.  In this experiment, subjects first
tapped along with an externally provided beat (synchronization), and
then continued tapping at the same speed without external guidance
(continuation).  During the continuation part, longer intervals were
associated with higher beta power during the interval, and vice versa.

\citeA{kononowicz_beta_2014} extended these findings to an interval
production task in humans, in which participants were instructed to
estimate time intervals of 2.5 seconds by pressing a key twice.  In
this study, intervals were distinct from each other, providing a clear
association between the beta power measured during the interval and
the length of the interval itself, and also provided enough time for
beta oscillations to fully resynchronize between trials.  As in
\citeauthor{bartolo_information_2014}'s study, beta power was
postitively correlated with the length of the produced intervals.

In another area of the field of timing, a large body of research
exists that links interval timing to reaction time tasks
\cite<e.g.,>{karlin_reaction_1959, drazin_effects_1961,
  grosjean_timing_2001}.  When participants come to expect the trigger
stimulus after a specific time interval, deviations from this interval
lead to predictable changes in reaction time.  For example,
\citeA{grosjean_timing_2001} found that when the stimulus appears
earlier than expected, reaction times will be slower.  Conversely,
when it appears later than expected, reaction times will be faster.  A
simple explanation for this effect is that participants build up the
readiness to react towards the point in time when they expect the
stimulus to occur.  When it appears before they expect it, they are
less ready to respond to it, and thus slower.  However, when it
appears later than anticipated, the readiness is very high and the
reaction comes faster.

These findings clearly show that timing plays a role in reaction time
tasks, in that participants implicitly use their sense of time to
predict when the stimulus will occur.

The purpose of the present study is to investigate the link between
beta power, interval timing, and reaction time.  As beta power has
been found to increase the length of estimated time intervals
\cite<e.g.>{kononowicz_beta_2014}, and interval timing has been shown
to play a role in reaction time tasks via expectancy effects
\cite<e.g.>{grosjean_timing_2001}, we hypothesise that beta power can
be used to predict the expectancy, and hence the response speed, in a
reaction time task.

To assess this hypothesis, we performed a simple reaction time task in
which participants initiated each trial themselves by pressing a key.
Beta power was measured during the first second of each trial.  For
the first 150 trials, the trigger stimulus always occured after a 1.8
s from the start of the trial, in order for subjects to come to expect
the stimulus after this interval.  In the last 250 trials, the
interval was chosen at random from the standard interval, and both 100
ms longer and shorter intervals.  The interval lengths were chosen
such that enough time passed between start of the trial and stimulus
onset to allow the beta power to build up to its peak within the
interval \cite<see>{kononowicz_beta_2014}.

In line with the findings of \citeA{grosjean_timing_2001}, reaction
times were hypothesized to be slower for the shorter interval, and
faster on trials with the longer interval.

In addition, we expected reaction times to be faster in trials where
beta power is low, because participants hypothetically estimate the
interval to end sooner than it actually does, and are thus very ready
to respond when the stimulus finally does occur.  Conversely, in
trials with high beta power, participants should estimate the stimulus
to occur later than it does, and thus not be prepared for it when it
actually occurs, leading to a slower reaction time.

\section{Method}
\section{Results}

\begin{table}[ht]
  \caption{Regression of reaction time on length condition and beta power}
  \centering
  \begin{tabular}{lrrrr}
    \toprule
                                                  & B        & SE         & t       & p         \\
    \midrule
    Intercept                                     & -0.0029  & 0.0008     &  -3.42  & $< 0.001$ \\
    Length condition: long                        & -0.0171  & 0.0016     & -10.44  & $< 0.001$ \\
    Length condition: short                       &  0.0259  & 0.0015     &  16.96  & $< 0.001$ \\
    Beta power                                    & -0.0031  & 0.0063     &  -0.49  & $0.622$   \\
    Beta power $\times$ Length condition: long    & -0.0079  & 0.0123     &  -0.64  & $0.521$   \\
    Beta power $\times$ Length condition: short   & -0.0086  & 0.0130     &   0.66  & $0.511$   \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[ht]
  \caption{Descriptive statistics overall as well as by length condition}
  \centering
  \begin{tabular}{lrrrrr}
    \toprule
                             & \multicolumn{2}{l}{Reaction time}              && \multicolumn{2}{l}{Beta}                       \\ \cmidrule{2-3} \cmidrule{5-6}
                             & \multicolumn{1}{l}{M} & \multicolumn{1}{l}{SD} && \multicolumn{1}{l}{M} & \multicolumn{1}{l}{SD} \\
    \midrule
    Overall                  & 0.188                 &  0.067                 && 0.277                 & 0.188                  \\
    Length condition: short  & 0.207                 &  0.070                 && 0.270                 & 0.168                  \\
    Length condition: medium & 0.188                 &  0.061                 && 0.283                 & 0.200                  \\
    Length condition: long   & 0.166                 &  0.070                 && 0.271                 & 0.177                  \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Discussion}

%%%%% Points to include
% solid effect of interval length
%   timing plays a role in RT
%     grosjean, karlin, drazin replicated/extended
%     -> robust effect
%
% no effect of beta
%   implicit vs explicit timing
%   internal vs external cues
%     coull & nobre 2008
%       Implicit timing; Exogenous temporal expectations
%         parietal cortex and cerebellum
%       Explicit timing; motor timing
%         Basal ganglia -> where joundi 2013 found beta suppression during tapping

% !! coull 2008 -> explicit/implicit different
% coull 2011 -> internal/external same
%
% ! n채채t채채nen hazard function (foreperiod etc)

% Hazard function: conditional probability that a target will occur at a particular time, given that it has not already occurred (Coull 2008)

% beta -> motor inhibition:
%   kononowicz: S can plan when to make second press -> beta at start
%   we:         S has to wait for beep -> no motor suppression at start

In the present study we investigated the connection between interval
timing, beta power, and reaction time. We tested whether presenting
the stimulus to react to earlier or later than expected would
influence response speed, and given this effect, whether beta activity
would influence reaction time by changing the time at which
participants expect the stimulus.

\subsection{Interval Length}
The first of our hypotheses, predicting that reaction times will be
slower in trials with a short interval, and faster in trials with a
long interval, was clearly supported by the data.  With the short and
long intervals leading to a 25 ms and a 17 ms difference in reaction
time to the standard condition, respectively, this effect is very
substantial given that the difference between the conditions was only
100 ms.

As expected from previous studies \cite<e.g.,>{karlin_reaction_1959,
  drazin_effects_1961, grosjean_timing_2001}, participants reacted
slower in trials with short intervals, and faster in trials with long
intervals. The decrease in response speed for short intervals can be
accounted for by the fact that participants are likely not yet
prepared to react, as they don't expect the stimulus to occur at this
time.

The faster response speed on trials with long intervals is more
interesting. It would seem reasonable to expect participants to
increase their preparedness to react until the point when they expect
the stimulus, and then either stay on that level of preparedness, or
even ``lose interest'' and become slower again. Our findings, however,
suggest that preparedness increases even further after the point in
time when the stimulus is expected to occur. This supports research by
\citeA{naatanen_diminishing_1970}, who argued that the increase in
response speed in fact stems from the passage of time. As participants
know that the stimulus will occur eventually, the more time passes,
the likelier it becomes that it will occur in the next moment. This
perspective leads to the prediction that reaction time becomes faster
the more time elapses before the stimulus, which is in fact what we
found. It would be an intriguing question for future research whether
this effect holds over a wider timespan.

\subsection{Beta Power}
Our second hypothesis, that beta power will correlate positively with
reaction time, was not confirmed.  In fact, Bayesian analyses even
suggest that it does not have any value as a predictor of reaction
time.

This finding can be interpreted using the taxonomy proposed by
\citeA{coull_dissociating_2008}.  \citeA{kononowicz_beta_2014}'s task
required motor timing, a form of explicit timing that has been shown
to be associated with activation of the basal ganglia
\cite{coull_dissociating_2008}.  The present study, in contrast, uses
a task involving implicit timing; more specifically, exogenous
temporal expectations, which involve multiple different brain areas,
but not the basal ganglia \cite{coull_dissociating_2008}.  As the beta
power that is associated with timing functions has been described as
originating from the basal ganglia, and only in studies employing
explicit timing tasks \cite<e.g.>{joundi_persistent_2013,
  bartolo_information_2014}, it is plausible that beta power does not
index the implicit timing required for the present study.  Thus, our
finding that beta power is not suited as a predictor of reaction time
reinforces the idea proposed by \citeA{coull_dissociating_2008} that
implicit and explicit timing employ different neural structures.

Another possible reason for beta power not being useful to predict
reaction time is that our task differed from those of
\citeA{kononowicz_beta_2014}, \citeA{bartolo_information_2014}, and
\citeA{joundi_persistent_2013} in that it did not allow participants
to fully plan when to make the second key press directly after the
first one, as they had to wait for the stimulus to respond. In other
words, while previous tasks required responses based purely on
internal cues (interval timing), the present task required
participants to wait for an external cue to respond. Thus, our finding
might suggest that internally and externally cued tasks employ
different timing mechanisms.

\subsection{Limitations}
The null findings might also be due to limitations of the study. The
room in which the experiment was conducted was not electrically
shielded, resulting in a lot of noise in the EEG signal. Furthermore,
we did not use the FCz electrode as \citeA{kononowicz_beta_2014} did,
but electrodes surrounding it. However, post-hoc analyses revealed a
significant relationship between subjects' mean beta power and mean
reaction time, indicating that beta activity was measured in a meaningful
way.

Moreover, our data did not satisfy all assumptions of the linear
regression model; normality, independence, and linearity were
violated.

\subsection{Brain Computer Interface}
A technical mistake in the implementation of the beta power
calculation led to a longterm drift in the beta values, so that the
BCI algorithm categorized almost all trials as either high or
low. Thus, the BCI did not provide the intended benefit of obtaining
as many non-standard interval lengths on trials with non-average beta
values as possible. Instead, length conditions were distributed
randomly without regard for the beta power in the current trial.

Given the potential benefit of a more beneficial distribution of
conditions, combined with no negative effects in case of failure, the
kind of BCI used in the present study can safely be used in future
research of a similar nature.

\subsection{Conclusion}


%%%%% Hedderik to Simon

% no separate future research section; include it into rest of text if at all

% talk about both internal/external cues & explicit/implicit timing

% argue with motor readiness about internal/external (for hedderik's experiment: participants can already plan when to press second time directly after the first press)
% argue with different brain regions about explicit/implicit

% explicit timing seems to be needed for motor planning

% talk about bci

% small conclusion in the end is nice
% last sentence should be positive

% no limitations section (findings in foreground, mention limitations only in context of findings)

\bibliographystyle{apacite} % / apalike
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/apacite/apacite.pdf
\bibliography{/home/max/Dropbox/Psychology/Thesis/latex/clean_references}


\end{document}

\end{document}
